---
title: 'Dispatching S3 Methods for Quick Data Plots '
author: "Nick Salkowski"
date: "Thursday, March 26, 2015"
output:
  ioslides_presentation:
    keep_md: yes
---

## Class Systems

R is an object oriented language -- one with different
systems of classes!

S3 is just one of those class systems.

 - Some sloppiness is allowed in S3
     - So, it is fairly easy to learn to use (or abuse)
 - S3 is used for a lot of common R functions

```{r, echo = FALSE, results = 'hide'}
set.seed(1234)

knitr::opts_knit$set(tidy = FALSE)

suppressWarnings(suppressPackageStartupMessages(
  library("ggplot2")))
suppressWarnings(suppressPackageStartupMessages(
  library("dplyr")))
library(pithr)
library(magrittr)
```

## S3 Example

```{r}
X_fact <- factor(
  LETTERS[sample.int(5, size = 50, replace = TRUE)])
X_num <- rnorm(50, mean = 10, sd = sqrt(10))
summary(X_fact)
summary(X_num)
```

## How does that work?

The `summary` function has different methods, depending on the class of object it is given.

```{r}
summary
```

## Methods

```{r}
methods(summary)
```

## S3 Method Function

```{r}
summary.factor
```

## Plot is another example

```{r}
methods(plot)
```

## Here are some more examples

```{r}
methods(coef)
methods(hist)
```

## Creating Your Own Class

Recall, S3 is sort of sloppy.  You can just create an object and then tell R which class (or classes) it belongs to!
```{r}
X_fish <- structure(rpois(50, lambda = 5),
  class = c("fish", "integer"))
summary.fish <- function(object, ...) {
  fish <- object
  table(fish)
  }
summary(X_fish)
```

## Motivation

Sometimes it is nice to have a plot that summarizes 
your data.

 - Exploring
 - Debugging

There are lots of nice plotting functions, but they 
don't all work with all kinds of data.

```{r}
X1 <- c(rnorm(70), rep(c(-Inf, Inf), 10),
        rep(c(NaN, NA), 5))
X2 <- factor(c(
  LETTERS[sample.int(5, size = 70, replace = TRUE)],
  rep(NA, 30)))
```

## Histograms? (Part 1)

```{r}
hist(X1)
```

## Histograms? (Part 2)

```{r}
(try(hist(X2)))
```

## Frequency Plots? (Part 1)

```{r}
barplot(table(X2))
```

## Frequency Plots? (Part 2)

```{r}
barplot(table(X1))
```

## ggplot2? (Part 1)

```{r}
qplot(X1, binwidth = 0.5)
```

## ggplot2? (Part 2)

```{r}
qplot(X2)
```

## Drawbacks

 - `hist` only handles numbers and dates
 - `barplot` + `table` doesn't handle doubles well
 - No support for lists or data.frames
 - Non-Finite and Missing Values are often dropped
 
## So, I wrote a package . . .

 - I'm relatively new to writing packages 
 (So, it's a little rough around the edges)
 - It's available on GitHub
 - It's called **pithr**
 
## What is a *pith*?

 - One definition of *pith* is *the essence of something*.
 - When used as a verb, *pith* means *to remove the pith from*.  

In the context of the pithr package, a *pith* is a simple summary plot of the data contained in an object.  

That is, the *pith* of your data is the *essence* of your data.

## Installing the pithr package

pithr is not available on CRAN, but it is available on GitHub.  To install 
pithr directly from GitHub, first install the [devtools](https://github.com/hadley/devtools) package, then use `devtools::install_github`.

```
install.packages("devtools")
devtools::install_github("NickSalkowski/pithr")
```

## Using pithr

The most basic command in the pithr package is `pith`.  

`pith` handles a wide variety of data types and structures, producing frequency plots or histograms, depending on the data itself.

## Integers (Range <= 25)
```{r}
X_int <- rpois(n = 100, lambda = 5)
pith(X_int)
```

## Integers (Range > 25)
```{r}
X_int2 <- rpois(n = 100, lambda = 500)
pith(X_int2)
```

## Doubles
```{r}
X_num <- rnorm(n = 100, mean = 5, sd = sqrt(5))
pith(X_num)
```

## Characters
```{r}
X_char <- LETTERS[X_int]
pith(X_char)
```

## Factors
```{r}
X_fact <- factor(X_char)
pith(X_fact)
```

## Logicals
```{r}
X_log <- as.logical(rbinom(100, 1, 0.625))
pith(X_log)
```

## Dates
```{r, fig.height = 4}
X_Date <- as.Date(sample.int(10000, size = 100, replace = TRUE), 
  origin = "1970-01-01")
pith(X_Date)
```

## Matrices & Arrays
Matrices and arrays are collapsed into vectors by `pith`.  
```{r, fig.height = 4}
X_mat <- matrix(rpois(400, 100),  ncol = 20)
pith(X_mat)
```

## Lists and data.frames
If `pith` is given a list or data.frame, each element or column is summarized separately.

```{r, fig.height = 4}
par(mfrow = c(1,2))
pith(cars)
```

## NA Values

`pith` seperates NA values, and presents their frequencies separately

```{r, fig.height = 4}
pith(X2)
```

## Non-Finite Values

`pith` separates non-finite values, and presents their frequencies separately.
```{r, fig.height = 4}
pith(X1)
```

## Arguments

`pith` accepts several arguments that help control the plot.

- plot If TRUE, a plot is produced.
- xname The "name" of the data.  This is used to produce the main plot title.
- breaks, include.lowest, right These arguments are passed to the `hist` 
  function, if a histogram is produced.
- ... Additional plot arguments.

## Example with arguments

```{r}
pith(X1, xname = "Numeric Vector Example", breaks = 5, las = 1)
```

## pithy Returns

`pith` invisibly returns a pith class object, which is a list that contains summary statistics for the data.  

There are some situations where it would 
be useful to return the data object inself, instead. 

`pithy` is a convenience 
function that calls `pith`, but returns the object.

This behavior is particularly handy when used with the 
[magrittr](https://github.com/smbache/magrittr) and 
[dplyr](https://github.com/hadley/dplyr) packages.

## pithy Example
```{r}
X_unif <- pithy(runif(100))
```

## magrittr + pithy

```{r}
X_gamma <- rgamma(100, 2, 2) %>%
  pithy
```

## dplyr + pithy
```{r, fig.height = 4}
par(mfrow = c(1, 2))
iris %>% select(Sepal.Length) %>% pithy %>%
  transmute(SLsq = Sepal.Length ^ 2) %>% pith
```

## Special dplyr + pithy functions
pithr also contains helper functions `filter_pithy`, `select_pithy`, 
`mutate_pithy`, and `transmute_pithy` that filter, select, mutate, and 
transmute data sets before calling `pithy`.  

Just like `pithy`, they 
return the original data set (before any filtering, selecting, or mutating).

## select_pithy example
```{r, fig.height = 2.5}
par(mfrow = c(1, 2))
iris %>% select_pithy(Petal.Length) %>%
  transmute_pithy(SLsq = Sepal.Length ^ 2) %>% tbl_df
```

## Effort vs. Quality

I tend to think about plotting applications as either **artisinal** or **industrial**

 - **artisinal plots** are created by a skilled plotmaker for a specific purpose.  Quality is highly valued.
 - **industrial plots** usually aren't quite as good, but can be mass produced with much less effort per plot.
 
`pithr` is built to produce *useful* plots with minimal effort, so I consider definitely it industrial.

If you want to produce a small number of beautiful artisinal plots, you are almost certainly better off using `ggplot2`.

## Live Demo

